---
title: "Accessibility impact of transport infrastructure: Spatial assessment of Bogota's future metro system"
author: "Andrés Restrepo Jiménez"
date: '2023-06-19'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(r5r)
library(accessibility)
library(ggplot2)
library(aopdata)
library(gtfstools)
library(readr)
library(dplyr)
library(tidyr)
library(st)
library(httr)
library(sf)
library(patchwork)
library(tmap)
library(tmaptools)
library(forcats)
library(scales)
library(writexl)
library(DataExplorer)
library(osmextract)
library(osmdata)
library(leaflet) # for making the interactive map
library(elevatr) # DEM (Digital Elevation Model)
library(raster)
#library(terra) # Not installed
library(h3jsr) # H3 hegagonal grid
library(remotes)
library(h3)
```

## Temporary directory

Location

```{r}
tempdir()
```

Deletes

```{r}
# unlink(tempdir(), recursive = TRUE)
```

## Memory setup

```{r}
options(java.parameters = "-Xmx6G")
```

# Bus Rapid Transit network (BRT)

## Downloading the data (turn off)

Setting the temp file path

```{r}
# brt_path <- tempfile("brt", fileext = ".zip")
```

Downloading the BRT GTFS data

```{r}
# httr::GET(
#   "https://storage.googleapis.com/gtfs-estaticos/GTFS-2023-06-21.zip",
#   httr::write_disk(brt_path)
# )
```

Initial: The GTFS data is a static feed (2021-07-04) provided by the Transport Secretary of Bogota City Council available [here](https://datos.movilidadbogota.gov.co/datasets/599751603a494051b15ac3d4f1f38ec5/about).

Update: The data was updated with static feed (2023-06-21) provided by Transmilenio in their open data webpage and available [here](https://datosabiertos-transmilenio.hub.arcgis.com/documents/gtfs-est%C3%A1ticos-2023-06-21/about).



Review of tables of downloaded data

```{r}
# unzip(brt_path, list = TRUE)
```


```{r}
# brt_gtfs <- read_gtfs(brt_path)
# 
# names(brt_gtfs)
```
## Reading data locally

Local path

```{r}
local_path <- "Data/Raw/Transport/BRT/GTFS-2023-06-21.zip"
```

Reading data locally

```{r}
brt_gtfs <- read_gtfs(local_path)

names(brt_gtfs)
```

## Data inspection

### Profiling

```{r}
plot_str(brt_gtfs)
```

### Agency

```{r}
head(brt_gtfs$agency)
```

### Routes

```{r}
head(brt_gtfs$routes)
```

#### Id

Number of rows in route table

```{r}
nrow(brt_gtfs$routes)
```

Number of route id's.

```{r}
n_distinct(brt_gtfs$routes$route_id)
```

There was 547 different routes in the BRT network, and 1086 in the new data.

#### Description

Number of descriptions

```{r}
n_distinct(brt_gtfs$routes$route_desc)
```

Description texts

```{r}
unique(brt_gtfs$routes$route_desc)
```
Filtering route id by desc value

##### Main corredor

```{r}
route_id_main <- filter(brt_gtfs$routes,  route_desc == "TRONCAL") %>%
  dplyr::select(route_id)
```

Subset GTFS data by route id

```{r}
brt_gtfs_main <- filter_by_route_id(brt_gtfs, route_id_main$route_id, keep = TRUE)
```

Geometry of subsetted routes

```{r}
geo_shapes_main <- gtfstools::get_trip_geometry(brt_gtfs_main, file = "shapes")
```


##### Urban

```{r}
route_id_urban <- filter(brt_gtfs$routes,  route_desc == "URBANO") %>%
  dplyr::select(route_id)
```

Subset GTFS data by route id

```{r}
brt_gtfs_urban <- filter_by_route_id(brt_gtfs, route_id_urban$route_id, keep = TRUE)
```

Geometry of subsetted routes

```{r}
geo_shapes_urban <- gtfstools::get_trip_geometry(brt_gtfs_urban, file = "shapes")
```

##### Feeder

```{r}
route_id_feeder <- filter(brt_gtfs$routes,  route_desc == "ALIMENTADOR") %>%
  dplyr::select(route_id)
```

Subset GTFS data by route id

```{r}
brt_gtfs_feeder <- filter_by_route_id(brt_gtfs, route_id_feeder$route_id, keep = TRUE)
```

Geometry of subsetted routes

```{r}
geo_shapes_feeder <- gtfstools::get_trip_geometry(brt_gtfs_feeder, file = "shapes")
```

##### Feeder_V

```{r}
route_id_feeder_v <- filter(brt_gtfs$routes,  route_desc == "ALIMENTADOR_V") %>%
  dplyr::select(route_id)
```

Subset GTFS data by route id

```{r}
brt_gtfs_feeder_v <- filter_by_route_id(brt_gtfs, route_id_feeder_v$route_id, keep = TRUE)
```

Geometry of subsetted routes

```{r}
geo_shapes_feeder_v <- gtfstools::get_trip_geometry(brt_gtfs_feeder_v, file = "shapes")
```


##### Complementary

```{r}
route_id_comp <- filter(brt_gtfs$routes,  route_desc == "COMPLEMENTARIO") %>%
  dplyr::select(route_id)
```

Subset GTFS data by route id

```{r}
brt_gtfs_comp <- filter_by_route_id(brt_gtfs, route_id_comp$route_id, keep = TRUE)
```

Geometry of subsetted routes

```{r}
geo_shapes_comp <- gtfstools::get_trip_geometry(brt_gtfs_comp, file = "shapes")
```

##### Special

```{r}
route_id_special <- filter(brt_gtfs$routes,  route_desc == "ESPECIAL") %>%
  dplyr::select(route_id)
```

Subset GTFS data by route id

```{r}
brt_gtfs_special <- filter_by_route_id(brt_gtfs, route_id_special$route_id, keep = TRUE)
```

Geometry of subsetted routes

```{r}
geo_shapes_special <- gtfstools::get_trip_geometry(brt_gtfs_special, file = "shapes")
```

##### Saving 

```{r}
# # Main
# geo_shapes_main %>%
#   st_write("Data/Preprocessed/BRT/Network/BRT_Network.gpkg",
#            "Main",
#            delete_layer=TRUE)
# 
# geo_shapes_urban %>%
#   st_write("Data/Preprocessed/BRT/Network/BRT_Network.gpkg",
#            "Urban",
#            delete_layer=TRUE)
# 
# geo_shapes_feeder %>%
#   st_write("Data/Preprocessed/BRT/Network/BRT_Network.gpkg",
#            "Feeder",
#            delete_layer=TRUE)
# 
# geo_shapes_feeder_v %>%
#   st_write("Data/Preprocessed/BRT/Network/BRT_Network.gpkg",
#             "Feeder_v",
#             delete_layer=TRUE)
# 
# geo_shapes_comp %>%
#   st_write("Data/Preprocessed/BRT/Network/BRT_Network.gpkg",
#            "Complementary",
#            delete_layer=TRUE)
# 
# geo_shapes_special %>%
#   st_write("Data/Preprocessed/BRT/Network/BRT_Network.gpkg",
#            "Special",
#            delete_layer=TRUE)




```


Test


```{r}
# # Main
# geo_shapes_main %>%
#   st_write("Data/Preprocessed/BRT/Network/BRT_Network.gpkg",
#            "Main")
# 
# geo_shapes_urban %>%
#   st_write("Data/Preprocessed/BRT/Network/BRT_Network.gpkg",
#            "Urban")
# 
# geo_shapes_feeder %>%
#   st_write("Data/Preprocessed/BRT/Network/BRT_Network.gpkg",
#            "Feeder")
# 
# geo_shapes_feeder_v %>%
#   st_write("Data/Preprocessed/BRT/Network/BRT_Network.gpkg",
#            "Feeder_v")
# 
# geo_shapes_comp %>%
#   st_write("Data/Preprocessed/BRT/Network/BRT_Network.gpkg",
#            "Complementary")
# 
# geo_shapes_special %>%
#   st_write("Data/Preprocessed/BRT/Network/BRT_Network.gpkg",
#            "Special")


```



#### Type

Number of types

```{r}
n_distinct(brt_gtfs$routes$route_type)
```

Type texts

```{r}
unique(brt_gtfs$routes$route_type)
```

Consistent with official GTFS [specification](https://gtfs.org/schedule/reference/#routestxt).

### Trips

```{r}
head(brt_gtfs$trips)
```

**How can you tell the direction?** **It does not have the shape id**

Number of routes in table

```{r}
n_distinct(brt_gtfs$trips$route_id)
```

Number of uniques entries

```{r}
n_distinct(brt_gtfs$trips)
```

There were 158976 trips included in the initial BRT GTFS data and 179454 in the updated data.

### Stops

```{r}
head(brt_gtfs$stops)
```

Number of stations

```{r}
n_distinct(brt_gtfs$stops)
```

Convert stops to sf

```{r}
brt_stations_sf <- st_as_sf(brt_gtfs$stops,coords = c("stop_lon", "stop_lat"), crs = 4326 )
```

Plot stations

```{r}
brt_stations_sf %>%
  st_geometry() %>%
  plot()
```

Basic geo map

```{r}
ggplot() +
  geom_sf(data = brt_stations_sf) +
  theme_minimal()
```

#### Id

Number of rows in station table

```{r}
n_distinct(brt_gtfs$stops$stop_id)
```

There were 7819 different stations in the BRT network, and 8214 in the new data.

#### Stop code

Number of rows in station table

```{r}
n_distinct(brt_gtfs$stops$stop_code)
```

Stop code and stop id seem to be equivalent.

### Calendar

```{r}
head(brt_gtfs$calendar)
```

From 1-5, it is the same schedule in the intial data, 2-5 in the new data.


### Calendar dates

```{r}
head(brt_gtfs$calendar_dates)
```

Number of exceptions

```{r}
n_distinct(brt_gtfs$calendar_dates)
```

#### Exception type

Exeption_type: "Indicates whether service is available on the date specified in the date field.

Valid options are:

1 - Service has been added for the specified date. 2 - Service has been removed for the specified date."

```{r}
unique(brt_gtfs$calendar_dates$exception_type)
```

Number of different dates in calendar dates

```{r}
unique(brt_gtfs$calendar_dates$date)
```

It seems to be holidays in Colombia.

#### Service id

```{r}
n_distinct(brt_gtfs$calendar_dates$service_id)
```




### Shapes

```{r}
head(brt_gtfs$shapes)
```

Number of shapes entries in table

```{r}
n_distinct(brt_gtfs$shapes$shape_id)
```

Number of unique entries

```{r}
n_distinct(brt_gtfs$shapes)
```

#### Geometry

Getting geometry from new data

```{r}
geo_shapes <- gtfstools::get_trip_geometry(brt_gtfs, file = "shapes")
```

The new object is already a sf object.

Review

```{r}
head(geo_shapes)
```

###### Saving

Saving shapes as geopackage (not working)

```{r}
# geo_shapes %>%
#   st_write("Data/Preprocessed/BRT/Network/BRT_Network.gpkg",
#            "Complete_Network",
#            delete_layer=TRUE)
```

Plotting geometry onlyv (too big)

```{r}
# geo_shapes %>%
#   st_geometry() %>%
#   plot()
```

Basic geo map (too big)

```{r}
# ggplot() +
#   geom_sf(data = geo_shapes) +
#   theme_minimal()
```

##### Sample of geometries

```{r}
set.seed(123) 

sample <- sample(brt_gtfs$trips$trip_id,1000)

head(sample)
```
Subsetting geometries

```{r}
geo_shapes_sample<- gtfstools::get_trip_geometry(brt_gtfs,file = "shapes", trip_id = sample)
```


```{r}
head(geo_shapes_sample)
```
Plotting sample geometries

```{r}
ggplot() +
  geom_sf(data = geo_shapes_sample) +
  theme_minimal()
```


##### Check for valid geometries

```{r}
unique(st_is_valid(geo_shapes))
```
There are invalid geometries in the new data.

##### Empty geometries (not working)

```{r}
unique(st_is_empty(geo_shapes))
```
There are not empty geometries in new data set.

##### Remove invalid geometries (not used in new data yet)

```{r}
# invalid_geometries <- !st_is_valid(geo_from_stops_filtered)
# 
# geo_from_stops_filtered <- geo_from_stops_filtered[invalid_geometries == FALSE, ]
```


### Stops times

```{r}
summary(brt_gtfs$stop_times)
```

```{r}
head(brt_gtfs$stop_times)
```

```{r}
brt_gtfs$stop_times
```

Filter stop sequence 0 values

```{r}
filter(brt_gtfs$stop_times, stop_sequence == 0)
```

Initial data: Count of 0 stop sequence values. New data: Zero entries with 0 value in stop sequence field.

```{r}
stop_cero_sequence <- filter(brt_gtfs$stop_times, stop_sequence == 0) %>%
  group_by(stop_id) %>%
  summarize(count_cero = n()) #%>%

stop_cero_sequence <- stop_cero_sequence[order(desc(stop_cero_sequence$count_cero)),]

```

Blank output.

#### Trip Id

```{r}
n_distinct(brt_gtfs$stop_times$trip_id)
```

Sorting by trip id

```{r}
brt_gtfs$stop_times[order(trip_id)]
```

**Review suffix structure**

#### Stop sequence

Unique value in stop_sequence field

```{r}
n_distinct(brt_gtfs$stop_times$stop_sequence)
```

Histogram on stop sequence values

```{r}
histo_stop_sequence <- ggplot(brt_gtfs$stop_times, aes(x=stop_sequence)) +
  ggtitle("Histogram of stop sequence values") + 
  geom_histogram() + xlab("Stop sequence")
histo_stop_sequence
```

Multiple stations with "1" values in the stop sequence.

Filter entries with zero values in stop time stop sequence field

```{r}
filter(brt_gtfs$stop_times, stop_sequence==0)
```

Zero entries with zero value in the stop sequence field

Filter a specific trip

##### Number of stations by trip (remove)


Adjusted count

```{r}
# stops_by_trip <- brt_gtfs$stop_times %>%
#   group_by(trip_id, stop_sequence) %>%
#   summarize(count_stop_sequence = n()) %>%
#   pivot_wider(names_from = stop_sequence, values_from = count_stop_sequence, values_fill = 0,names_prefix = "S_")
# 
# head(stops_by_trip)
```
Extract stops_by_trip to excel

```{r}
# write_xlsx(stops_by_trip,"Test/stops_by_trip.xlsx")
```



##### Case 1 

Two stops

```{r}
filter(brt_gtfs$stop_times, trip_id == "BC42D0021-04-2_T_1785")
```

Review of individual geometry

```{r}
geo_case_1 <- gtfstools::get_trip_geometry(brt_gtfs, trip_id = "BC42D0021-04-2_T_1785")
```

The new object is already a sf object.

Plotting geometry only

```{r}
ggplot() +
  geom_sf(data = geo_case_1) +
  theme_minimal()
```

Although, there are only 2 points, with the shape table data, the geometry output shows a smooth route.

##### Case 2

Tree stops

```{r}
filter(brt_gtfs$stop_times, trip_id == "BO7ED0007-02-2_T_2192")
```

Review of individual geometry

```{r}
geo_case_2 <- gtfstools::get_trip_geometry(brt_gtfs, trip_id = "BO7ED0007-02-2_T_2192")
```


Plotting geometry only

```{r}
ggplot() +
  geom_sf(data = geo_case_2) +
  theme_minimal()
```

##### Case 3

Four stops

```{r}
filter(brt_gtfs$stop_times, trip_id == "BCGIA0007-13-3_Z_6441")
```

Same example with -2 suffix

```{r}
filter(brt_gtfs$stop_times, trip_id == "BCGIA0007-13-3_Z_6441-2")
```

Review of individual geometry

```{r}
geo_case_3 <- gtfstools::get_trip_geometry(brt_gtfs, trip_id = "BCGIA0007-13-3_Z_6441")
```


Plotting geometry only

```{r}
ggplot() +
  geom_sf(data = geo_case_3) +
  theme_minimal()
```
__Non-continous geometry__


### Frequencies (blank)

```{r}
head(brt_gtfs$frequencies)
```

```{r}
brt_gtfs$frequencies
```

There is no data in the frequencies table.

### Fare attributes

```{r}
head(brt_gtfs$fare_attributes)
```

```{r}
brt_gtfs$fare_attributes
```

Types of fare (price) information by agency.

Transfers
1. 0 indicate no transfers permitted.
NA. indicates Unlimited transfers are permitted.

### Fare rules

```{r}
head(brt_gtfs$fare_rules)
```

```{r}
brt_gtfs$fare_rules
```

#### Fare id

```{r}
unique(brt_gtfs$fare_rules$fare_id)
```
Fare 2 missing

#### Route id

```{r}
length(unique(brt_gtfs$fare_rules$route_id))
```
Fare information for 1086 routes.


### Route detailed review

#### Case 1

Route 6-4, "6-4" in route id field with id 

```{r}
brt_gtfs_C1 <- filter_by_route_id(brt_gtfs, "T_1899", keep = TRUE)
```


##### Route

```{r}
brt_gtfs_C1$routes
```
##### Trips

```{r}
brt_gtfs_C1$trips
```

Schedule: Monday to Friday.

Number of trips

```{r}
length(unique(brt_gtfs_C1$trips$trip_id))
```
There are 666 trips for this route.


##### Shapes

```{r}
brt_gtfs_C1$shapes
```

Geometry

```{r}
geo_brt_gtfs_C1 <- gtfstools::get_trip_geometry(brt_gtfs_C1,file = "shapes")
```

Review

```{r}
geo_brt_gtfs_C1
```

Plotting

```{r}
ggplot() +
  geom_sf(data = geo_brt_gtfs_C1) +
  theme_minimal()
```
##### Stops

```{r}
brt_gtfs_C1$stops
```

Convert stops to sf

```{r}
geo_brt_stations_C1_sf <- st_as_sf(brt_gtfs_C1$stops,coords = c("stop_lon", "stop_lat"), crs = 4326)
```

Plot stations

```{r}
geo_brt_stations_C1_sf %>%
  st_geometry() %>%
  plot()
```
##### Stop times

```{r}
arrange(brt_gtfs_C1$stop_times, trip_id, stop_sequence)
```


##### Fare attribute

```{r}
brt_gtfs_C1$fare_attributes
```

##### Fare rules

```{r}
brt_gtfs_C1$fare_rules
```
##### Plotting

```{r}
ggplot() +
  geom_sf(data = geo_brt_stations_C1_sf) + 
  geom_sf(data = geo_brt_gtfs_C1) + 
  theme_minimal()
```


## Data cleaning (pending)

# OSM network (pending)

## Data reading

Open Street Map Data downloaded from: <https://export.hotosm.org/es/v3/>

```{r}
osm_network <- oe_read("Data/Raw/Transport/OSM/Bogota_general_network.osm.pbf") 
```

Plotting

```{r}
par(mar = rep(0.1, 4))
plot(sf::st_geometry(osm_network))
```

## Data inspection

### CRS

```{r}
st_crs(osm_network)
```

### Profilling

```{r}
plot_missing(osm_network)
```

Filter by highway value

```{r}
osm_by_highway<- osm_network %>%
  group_by(highway) %>%
  summarize(count_entry = n()) %>%
  arrange(desc(count_entry))
  

osm_by_highway
```

Plotting by highway value

```{r}
osm_subsetting <- osm_by_highway %>%
  filter(., highway == "residential")


par(mar = rep(0.1, 4))
plot(sf::st_geometry(osm_subsetting))
```

Interactive map by subsetting value

```{r}
# leaflet() %>% 
#   addTiles() %>% 
#   addPolylines(data = osm_subsetting)

```

**What sub-setting should be done before the accessibility modelling?**

## Data cleaning

# Metro system (pending)

## GTFS data generation

Generates the main tables of the GTFS metro data.

### Agency

```{r}
metro_agency <- tibble::tribble(
  ~agency_id,  ~agency_name,        ~agency_url,  ~agency_timezone,  ~agency_lang,  ~agency_phone,
  "1", "Metro de Bogotá", "https://www.metrodebogota.gov.co/", "America/Bogota", "es","(+57)601-555-33-33")

```

Generates table

```{r}
data.table::setDT(metro_agency)
```


### Routes

Create a metro route table

```{r}
metro_routes <- tibble::tribble(
  ~route_id,  ~route_short_name,~route_long_name, ~route_desc, ~agency_id, ~route_color, ~route_text_color, ~route_type,
  "M_001", "S_001", "South", "Main", "1", "FFFFFF", "000000", "1",
  "M_002", "N_001", "North", "Main", "1", "FFFFFF", "000000", "1")
```

Reads it as data frame

```{r}
data.table::setDT(metro_routes)
```



### Trips

### Stops

Reading stations data of first line of metro.

```{r}
raw_metro_stations <- st_read("Data/Raw/Transport/Metro/Stations/ESTACIONES.shp") %>%
  st_transform(.,4326)
```
Review CRS

```{r}
st_crs(raw_metro_stations)
```
Geometry

```{r}
qtm(raw_metro_stations)
```



### Calendar

### Calendar dates

### Shapes

Reading track data of first line of metro.

```{r}
raw_metro_line <- st_read("Data/Raw/Transport/Metro/Track/TRAZADO_PLMB.shp") %>%
  st_transform(.,4326)
```

Review CRS

```{r}
st_crs(raw_metro_line)
```

Geometry

```{r}
qtm(raw_metro_line)
```
Create shape id column

```{r}
raw_metro_line$shape_id = "M1_0001"
```

Relocate columns

```{r}
raw_metro_line <- raw_metro_line  %>%
  relocate(shape_id)
```

Create shape table from sf object

```{r}
metro_line_shape <- convert_sf_to_shapes(raw_metro_line, shape_id = NULL)
```



### Stop times

### Frequencies (non-existing in BRT data)

### Fare attributes

### Fare rules


## Data inspection

## Data cleaning

# Multipurpose survey

## Reading the data

### ZPU layer

```{r}
geo_zpu <- st_read("Data/Raw/Demo/Survey/Shape/ShapeEM2021/EM2021_UPZ.shp")
```

Geometry

```{r}
qtm(geo_zpu)
```

### Survey data

#### Main forms

Downloading and reading from source

```{r}
# url<- "https://www.sdp.gov.co/sites/default/files/encuestas-multiproposito/20230620em2021.zip"
# 
# download.file(url,"Data/Survey/CSV_Main/20230620em2021.zip")
# 
# unzip("Data/Survey/CSV_Main/20230620em2021.zip",exdir = "Data/Survey/CSV_Main")
# 
# survey_raw <- read_csv("Data/Survey/CSV_Main/20230620EM2021.csv")
```

Reading from local

```{r}
survey_raw <- read_csv("Data/Raw/Demo/Survey/CSV_Main/20230620EM2021.csv")
```

Preview

```{r}
head(survey_raw,10)
```

#### Additional variables

Downloading and reading from source

```{r}
# url<- "https://www.sdp.gov.co/sites/default/files/encuestas-multiproposito/20230220_variables_adicionales_em2021_csv.zip"
# 
# download.file(url,"Data/Survey/CSV_Additional/20230220_variables_adicionales_em2021_csv.zip")
# 
# unzip("Data/Survey/CSV_Additional/20230220_variables_adicionales_em2021_csv.zip",exdir = "Data/Survey/CSV_Additional")
# 
# survey_add_raw <- read_csv("Data/Survey/CSV_Additional/20230220_variables_adicionales_em2021.csv")
```

Reading from local

```{r}
survey_add_raw <- read_csv("Data/Raw/Demo/Survey/CSV_Additional/20230220_variables_adicionales_em2021.csv")
```

Preview

```{r}
head(survey_add_raw,10)
```

## Data inspection

### ZPU layer

CRS

```{r}
st_crs(geo_zpu)
```

Summary

```{r}
summary(geo_zpu)
```

UPZEM: Numeric UPZEM2: String

### Survey data

#### Main forms

Data type

```{r}
Datatypelist_survey <- survey_raw %>% 
  summarise_all(class) %>%
  pivot_longer(everything(), 
               names_to="All_variables", 
               values_to="Variable_class")

Datatypelist_survey
```

##### Profiling

Summary

```{r}
introduce(survey_raw)
```

Missing values

```{r}
plot_missing(survey_raw)
```

Missing values on ID fields

```{r}
survey_raw_id <- survey_raw %>% 
  dplyr::select(directorio, directorio_per, dpto, mpio, clase, cod_localidad, nombre_localidad, cod_upz_grupo, nombre_upz_grupo, estrato2021, nombre_estrato, fex_c)

plot_missing(survey_raw_id)

```

Missing values by ZPU

```{r}
na_by_zpu <- survey_raw %>%
  group_by(cod_upz_grupo) %>%
  summarize(count_entry = n()) %>%
  mutate(percentage_entry = count_entry / nrow(survey_raw) * 100) %>%
  arrange(desc(percentage_entry))

head(na_by_zpu)
```

Missing ZPU by "directorio"

```{r}
entry_by_directorio <- survey_raw %>%
  group_by(directorio) %>%
  summarize(count_entry = n(),count_na = sum(is.na(cod_upz_grupo))) %>%
  mutate(percentage_entry = count_entry / nrow(survey_raw) * 100) %>%
  mutate(control = ifelse(count_entry == count_na,"All na's", "None")) %>%
  mutate(diff = count_entry - count_na) %>%
  arrange((diff))

entry_by_directorio
```

Filter equal values

```{r}
entry_by_directorio_control <- filter(entry_by_directorio, control == "None") %>%
  mutate(control_diff = diff==count_entry)

unique(entry_by_directorio_control$control_diff)
```

All entries, either have all NA's value per "directorio" or none.

Missing ZPU

```{r}
filter(survey_raw, is.na(cod_upz_grupo))

```

##### Example

directorio = "220102"

```{r}
filter(survey_raw, directorio == "220102") %>%
  dplyr::select(directorio,directorio_per, directorio_hog, # ID
         cod_upz_grupo, # ZPU code
         nombre_upz_grupo, # ZPU name
         orden, # Order in home
         npcep4, # Age
         npcep6, # Relation to Household leader
         npcep11a) # Place of birth
```

#### Additional variables

Data type

```{r}
Datatypelist_survey_add <- survey_add_raw %>% 
  summarise_all(class) %>%
  pivot_longer(everything(), 
               names_to="All_variables", 
               values_to="Variable_class")

Datatypelist_survey_add
```

##### Profiling

Summary

```{r}
introduce(survey_add_raw)
```

Missing values

```{r}
plot_missing(survey_raw)
```

Missing values on ID fields

```{r}
survey_add_raw %>% dplyr::select(directorio, directorio_per, directorio_hog) %>%
  plot_missing(.)

```

##### Example

directorio = "220102"

```{r}
filter(survey_add_raw, directorio == "220102") %>%
  dplyr::select(
    directorio,directorio_per, # ID
    N_fuerza_trabajo, # In labour market
    N_informal, # Informal
    N_analfabeta, # Illiterate
    N_pobre_ipm, # MPI poor
    N_codigo_upz_trabajo, # ZPU ID of workplace
    N_INGTOT_PER, # Total income 
    N_pobre_monetario) # Monetary poverty
```

# Land use (in progress)

## Data reading

```{r}
land_use_terrain <- st_read("Data/Raw/Land_use/Terrain/DestinoSHP/Destino.shp")
```
Geometry

```{r}
# qtm(land_use_terrain)
```

Invalid poligons

## Data inspection

CRS

```{r}
st_crs(land_use_terrain)
```

Set the polygon object in WGS84 CRS

```{r}
land_use_terrain84 <- land_use_terrain %>%
  st_set_crs(.,4326)
```

New CRS


```{r}
st_crs(land_use_terrain84)
```

## Data cleaning

Invalid geometries

```{r}
invalid_geo_land_use <- st_is_valid(land_use_terrain84, reason = TRUE)
```

Clean geometry

```{r}
land_use_terrain84_clean <- st_make_valid(land_use_terrain84)
```


Map

```{r}
qtm(land_use_terrain84_clean)
```





# Elevation data (Not included yet)

Elevation data on every ZPU polygon

**Should be calculated for the H3 hegagon grid**

```{r}
# elevation <- get_elev_raster(geo_zpu, z = 9)
```

```{r}
# elevation
```

Plotting test

```{r}
# plot(elevation)
# plot(geo_zpu, add = TRUE)
```

## Data cleaning (pending,remove?)

# Preprocessing

## Spatial analysis unit

H3 hexagonal grid

Set the polygon object in WGS84 CRS

```{r}
geo_zpu_wgs84 <- geo_zpu %>%
  st_transform(.,4326)
```

New CRS

```{r}
st_crs(geo_zpu_wgs84)
```
Dissolve polygons

Method 1

```{r}
geo_zpu_wgs84_mono <- st_union(geo_zpu_wgs84, by_feature = FALSE)
```

Method 2

```{r}
geo_zpu_wgs84_mono_2 <- geo_zpu_wgs84 %>%
  group_by() %>%
  summarise()
```



### Hegagonal grid

#### Resolution 9

Intersecting grids

```{r}
intergrid_9 <- polyfill(geo_zpu_wgs84_mono, res = 9)
```
H3 id to sf

```{r}
hex_grid_9<- h3_to_geo_boundary_sf(intergrid_9)
```

Basic geo map

```{r}
ggplot() +
  geom_sf(data = hex_grid_9) +
  theme_minimal()
```
There are holes in the grid with a 9 resolution.

#### Resolution 8

Intersecting grids

```{r}
intergrid_8 <- polyfill(geo_zpu_wgs84_mono, res = 8)
```
H3 id to sf

```{r}
hex_grid_8 <- h3_to_geo_boundary_sf(intergrid_8)
```

Basic geo map

```{r}
ggplot() +
  geom_sf(data = hex_grid_8) +
  theme_minimal()
```

There are not holes with 8 resolution value.



## Data enrichment


### Survey

#### Main forms

#### Additional variables (pending)

Aggregation of additional variables data by ZPU

```{r}
# survey_add_raw_clean <- survey_add_raw %>%
  
```

Erase entries with Na ZPU id field.

```{r}

```

### Land use

Spatial join between hexagon grid and land use layer

```{r}
# hex9_land_use_join <- st_join(hex_grid_9, land_use_terrain84_clean, join = st_intersects )
```



Pivot wide by land use category

```{r}
# hex9_land_use_join_wide <- hex9_land_use_join %>%
# 
#   group_by(h3_index, DESTINOCOD) %>%
#   summarize(count_land_use_code = n())  %>%
#   pivot_wider(names_from = "DESTINOCOD", values_from = "count_land_use_code", values_fill = 0 , names_prefix = "LU_COD_")
# 
# 
#   
```
Review

```{r}
# head(hex9_land_use_join_wide)
```
Dropping geometry column

```{r}
# hex9_land_use_join_wide <- hex9_land_use_join_wide %>%
#   dplyr::select(!geometry) %>%
#   st_drop_geometry(.)
```


Review

```{r}
# head(hex9_land_use_join_wide)
```

Merge with initial hexagon grid

```{r}
# hex9_land_use_final <- merge(hex_grid_9, hex9_land_use_join_wide, by.x = "h3_index", by.y = "h3_index", all.x = TRUE)

```

Saving hexagon grid

__It does not include demographics data yet__

```{r}
# hex9_land_use_final %>%
#   st_write("Data/Preprocessed/Grid/Hex9_grid.gpkg",
#            "Hex9_LU",
#            delete_layer=TRUE)
```

Reading from preprocessed directory

```{r}
hex9_land_use_final <- st_read("Data/Preprocessed/Grid/Hex9_grid.gpkg","Hex9_LU")
```


Review

```{r}
head(hex9_land_use_final)
```
Plotting

```{r}
tmap_mode("plot")
qtm(hex9_land_use_final, fill = "LU_COD_01")
```


## Origin and destination points

```{r}
hex9_points <- hex9_land_use_final %>%
  st_centroid() %>%
  rename(id = h3_index)
  
```

```{r}
qtm(hex9_points)
```

Calculate lon and lat coordinates

```{r}
hex9_points$lon <- st_coordinates(hex9_points)[, 1]
hex9_points$lat <- st_coordinates(hex9_points)[, 2]
```

Drop geometry of hex grid

```{r}
hex9_points <- hex9_points %>% 
  st_drop_geometry() %>%
  relocate(lon, .after = id) %>%
  relocate(lat, .after = lon)
```

Save point data as csv

```{r}
write.csv(hex9_points,file='Data/Preprocessed/Grid/Hex9_grid_LU.csv', row.names=FALSE)
```





# Spatial analysis

# Accessiblity modelling

## Data preparation

Creates directory for accessibility analysis


```{r}
access_dir <- "Accessibility"
dir.create(access_dir)
```

Creates before and after directories

```{r}
before_dir <- file.path(access_dir, "Before")
after_dir <- file.path(access_dir, "After")

dir.create(before_dir)
dir.create(after_dir)
```

Copy input files to before directory


```{r}
# BRT GTFS data
file.copy(from = "Data/Raw/Transport/BRT/GTFS-2023-06-21.zip", to = file.path(before_dir, "brt_gtfs.zip"))

# OSM Street network
file.copy(from = "Data/Raw/Transport/OSM/Bogota_general_network.osm.pbf", to = file.path(before_dir, "street_network.osm.pbf"))

# Hexagon opportunities data based on hexagon grid
file.copy(from = "Data/Preprocessed/Grid/Hex9_grid_LU.csv", to = file.path(before_dir, "hex9_points_LU.csv"))

```

Visualize directory

```{r}
fs::dir_tree(access_dir)
```



## Current scenario

It does not have the future metro data.

### Setup

```{r}
r5r_before <- setup_r5("Accessibility/Before", verbose = FALSE)
```
Centroids of hexagons

```{r}
hex_points_main <- data.table::fread("Accessibility/Before/hex9_points_LU.csv")
```

### Time travel matrix

Without erasing geometries

```{r}
ttm <- travel_time_matrix(
  r5r_before,
  origins = hex_points_main,
  destinations = hex_points_main,
  mode = c("WALK", "TRANSIT"),
  departure_datetime = as.POSIXct(
    "06-06-2023 8:00:00",
    format = "%d-%m-%Y %H:%M:%S"
  ),
  max_walk_time = 30,
  max_trip_duration = 120,
  verbose = FALSE,
  progress = FALSE
)
```

Rename columns of time travel output

```{r}
data.table::setnames(ttm, "travel_time_p50", "travel_time")
```


## Base scenario

### Accessibility package

#### Cumulative opportunities measure

High street retail opportunities

```{r}
cum_opportunities <- cumulative_cutoff(
  ttm,
  land_use_data = hex_points_main,
  opportunity = "LU_COD_21",
  travel_cost = "travel_time",
  cutoff = 30
)
```


Review

```{r}
head(cum_opportunities)
```

#### Minimum travel cost

To high street retail opportunities

```{r}
min_time <- cost_to_closest(
  ttm,
  land_use_data = hex_points_main,
  opportunity = "LU_COD_21",
  travel_cost = "travel_time"
)
```

Review

```{r}
head(min_time)
```

### R5R package

```{r}
r5r_access <- accessibility(
  r5r_before,
  origins = hex_points_main,
  destinations = hex_points_main,
  opportunities_colname = "LU_COD_21",
  decay_function = "step",
  cutoffs = 30,
  mode = c("WALK", "TRANSIT"),
  departure_datetime = as.POSIXct(
    "06-06-2023 8:00:00",
    format = "%d-%m-%Y %H:%M:%S"
  ),
  max_walk_time = 30,
  max_trip_duration = 120,
  verbose = FALSE,
  progress = FALSE
)
```

Review

```{r}
head(r5r_access)
```

### Results comparison

Using accessibility package

```{r}
cum_cutoff_29 <- cumulative_cutoff(
  ttm,
  land_use_data = hex_points_main,
  opportunity = "LU_COD_21",
  travel_cost = "travel_time",
  cutoff = 29
)
```

Merge results

```{r}
access_comparison <- merge(
  r5r_access,
  cum_cutoff_29,
  by = "id"
)
```


Setting columns names accross methods

```{r}
data.table::setnames(
  access_comparison,
  old = c("accessibility", "LU_COD_21"),
  new = c("r5r_access", "accessibility_access")
)
```

Review

```{r}
head(access_comparison[, .(id, r5r_access, accessibility_access)])
```
Results consistent across methods.

### Spatial distribution

Merging accessibility results with hexagon grid

```{r}
spatial_access <- merge(
  hex_grid_9,
  cum_opportunities,
  by.x = "h3_index",
  by.y = "id"
)
```

Plotting

Inferno Viridis color scheme

```{r}
ggplot(spatial_access) +
  geom_sf(aes(fill = LU_COD_21), color = NA) +
  scale_fill_viridis_c(option = "inferno") +
  labs(fill = "Accessible\nHigh Street Retail") +
  theme_minimal()
```


Regular Vidiris color scheme

```{r}
ggplot(spatial_access) +
  geom_sf(aes(fill = LU_COD_21), color = NA) +
  scale_fill_viridis_c() +
  labs(fill = "Accessible\nHigh Street Retail") +
  theme_minimal()
```

Save results in geopackage

```{r}
spatial_access %>%
  st_write("Data/Results/Before/Grid/Hex9_grid.gpkg",
           "Hex9_spatial_access",
           delete_layer=TRUE)
```
__Save in different directory__

Plotting of BRT Network

```{r}

```



## Alternative scenario
