---
title: "Accessibility impact of transport infrastructure: Spatial assessment of Bogota's future metro system"
author: "Andrés Restrepo Jiménez"
date: '2023-06-19'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(r5r)
library(accessibility)
library(ggplot2)
library(aopdata)
library(gtfstools)
library(dplyr)
library(st)
library(httr)
library(sf)
library(patchwork)
library(tmap)
library(forcats)
library(scales)
```

## Memory setup

```{r}
options(java.parameters = "-Xmx4G")
```

# Bus Rapid Transit network (BRT)

## Downloading the GTFS data

Setting the temp file path

```{r}
brt_path <- tempfile("brt", fileext = ".zip")
```

Downloading the BRT GTFS data

```{r}
httr::GET(
  "https://www.arcgis.com/sharing/rest/content/items/6def68cda70c43ab8d93d74f43c5915a/data",
  httr::write_disk(brt_path)
)
```
The GTFS data is a static feed (2021-07-04) provided by the Transport Secretary of Bogota City Council.

## Inspection of data

Review of tables of downloaded data

```{r}
unzip(brt_path, list = TRUE)
```

```{r}
brt_gtfs <- read_gtfs(brt_path)

names(brt_gtfs)
```
### Agency

```{r}
head(brt_gtfs$agency)
```


### Routes

```{r}
head(brt_gtfs$routes)
```
#### Id

Number of rows in route table

```{r}
nrow(brt_gtfs$routes)
```
Number of route id's.

```{r}
n_distinct(brt_gtfs$routes$route_id)
```
There are 547 different routes in the BRT network.

#### Description

Number of descriptions

```{r}
n_distinct(brt_gtfs$routes$route_desc)
```
Description texts

```{r}
unique(brt_gtfs$routes$route_desc)
```
#### Type

Number of types

```{r}
n_distinct(brt_gtfs$routes$route_type)
```
Type texts

```{r}
unique(brt_gtfs$routes$route_type)
```
Consistent with official GTFS [specification](https://gtfs.org/schedule/reference/#routestxt).

### Trips

```{r}
head(brt_gtfs$trips)
```
__How can you tell the direction?__
__It does not have the shape id__

Number of routes in table

```{r}
n_distinct(brt_gtfs$trips$route_id)
```
Number of duplicates

```{r}
brt_gtfs$trips[, .N, by = .(route_id, trip_id, service_id)]
```
Review

```{r}
n_distinct(brt_gtfs$trips)
```
There are 158976 trips included in the BRT GTFS data.

### Stops

```{r}
head(brt_gtfs$stops)
```
Number of stations

```{r}
n_distinct(brt_gtfs$stops)
```
#### Id

Number of rows in station table

```{r}
n_distinct(brt_gtfs$stops$stop_id)
```
There are 7819 different stations in the BRT network.

#### Stop code

Number of rows in station table

```{r}
n_distinct(brt_gtfs$stops$stop_code)
```
Stop code and stop id seem to be equivalent.

### Calendar

```{r}
head(brt_gtfs$calendar)
```
From 1-5, it is the same schedule.

### Shape (non existing)

__Do we need to create one? How?__

#### Geometry

Getting geometry from stops location for trip in case 3.

```{r}
geo_from_stops_test <- gtfstools::get_trip_geometry(brt_gtfs, trip_id = "BC26E0001-2-3")
```

Plotting

```{r}
plot(geo_from_stops_test)
```

Plotting geometry only


```{r}
geo_from_stops_test %>%
  st_geometry() %>%
  plot()
```

Basic geo map

```{r}
ggplot() +
  geom_sf(data = geo_from_stops_test) +
  theme_minimal()
```


#### Creating shape table


#### Converting to sf

```{r}
test_shape <- gtfstools::convert_shapes_to_sf(
  geo_from_stops,
  shape_id = "BC26E0001-2-3"
)
```
__Need to create a shape table before converting to sf__



### Stops times

```{r}
head(brt_gtfs$stop_times)
```
#### Trip Id


```{r}
n_distinct(brt_gtfs$stop_times$trip_id)
```

Sorting by trip id

```{r}
brt_gtfs$stop_times[order(trip_id)]
```
__How to interpret the stop_sequence field?__

Unique value in stop_sequence field

```{r}
n_distinct(brt_gtfs$stop_times$stop_sequence)
```
Filter a specific trip

Case 1 

```{r}
filter(brt_gtfs$stop_times, trip_id == "BC26E0001-10-3")
```
Case 2 

```{r}
filter(brt_gtfs$stop_times, trip_id == "BC26E0005-19-3")
```
__Do a group by with 0 and 1 values in stop sequence__


Case 3 

```{r}
filter(brt_gtfs$stop_times, trip_id == "BC26E0001-2-3")
```

### Frequencies (non existing)

### Calendar dates

```{r}
head(brt_gtfs$calendar_dates)
```
Number of exceptions

```{r}
n_distinct(brt_gtfs$calendar_dates)
```
#### Exception type

Exeption_type: "Indicates whether service is available on the date specified in the date field.

Valid options are:

1 - Service has been added for the specified date.
2 - Service has been removed for the specified date." 

```{r}
unique(brt_gtfs$calendar_dates$exception_type)
```
Number of different dates in calendar dates

```{r}
unique(brt_gtfs$calendar_dates$date)
```
It seems to be holidays in Colombia.

#### Service id

```{r}
n_distinct(brt_gtfs$calendar_dates$service_id)
```


